---
output: rmarkdown::github_document
---

# `covr` segfault reprex

A unique scenario, where a `fastmap` is created and used to store a `parquet`
`data.frame` produces a segfault when `covr` tries to save traces using
`saveRDS`.

In its simplest form, this would be produced from a package that tests a
function such:

```r
function(f) {
  m <- fastmap::fastmap()  
  m$set("data", arrow::read_parquet(f))
}
```

The details get far messier, so I've tested quite a few different scenarios
to try to pin down what is necessary to prompt a segfault:

## Parameters

* **fastmap value**: the type of value to be stored in the `fastmap`
* **fastmap init**: how the `fastmap` is initialized  
  either at
  [build, runtime, lazily](https://github.com/dgkf/reprex-covr-segfault/blob/02913e0e5d397a33574b056bc78cb168933aae69/R/parquet-fastmap.R#L1-L3),
  or in the [gloabl env](https://github.com/dgkf/reprex-covr-segfault/blob/02913e0e5d397a33574b056bc78cb168933aae69/R/onload.R#L5-L6)
* **`.onLoad` if()**: whether there is an `if` block in the `.onLoad` function  
  believe it or not, this affects whether `covr` segfaults in some cases.
* **covr source and version**: either released `covr` on **CRAN** or the dev
  build of covr.

Note that S3 methods are registered `onLoad` because registering them
using the `NAMESPACE` had a similar effect to using an `.onLoad` `if ()` block.

## Tests

<details>
<summary>
<strong><i>code</i></strong>
</summary>

```{r, warning = FALSE, message = FALSE}
# if a dev version of covr is installed in library, will need to be uninstalled

dir.create(lib_covr_cran <- file.path(tempdir(), "lib_covr_cran"))
install.packages("covr", lib = lib_covr_cran)

dir.create(lib_covr_dev <- file.path(tempdir(), "lib_covr_dev"))
remotes::install_github("r-lib/covr", lib = lib_covr_dev, force = TRUE, quiet = TRUE)
# pak::pkg_install("r-lib/covr", lib = lib_covr_dev)

scenarios <- expand.grid(
  "onLoad if()" = c(TRUE, FALSE),
  "covr remote" = c("covr", "r-lib/covr"),
  "covr version" = "",
  "fastmap init" = c("global", "lazy-stub", "lazy-fastmap", "runtime", "buildtime"),
  "fastmap value" = c("integer", "parquet"),
  "exit status" = 0,
  stringsAsFactors = FALSE
)

cols <- c("fastmap value", "fastmap init", "onLoad if()", "covr remote", "covr version", "exit status")
scenarios <- scenarios[, cols]

for (i in seq_len(nrow(scenarios))) {
  cat("running scenario", i, "... \n")
  scenario <- as.list(scenarios[i, ])

  onload <- readLines("R/onload.R")
  onload[[3]] <- if (scenario[["onLoad if()"]]) "  if (TRUE) {}" else ""
  writeLines(onload, "R/onload.R")

  envvars <- c(
    "REPREX_FASTMAP_INIT_STYLE" = scenario[["fastmap init"]],
    "REPREX_FASTMAP_VALUE_TYPE" = scenario[["fastmap value"]]
  )

  libs <- c(
    switch(scenario[["covr remote"]],
      "covr" = lib_covr_cran,
      "r-lib/covr" = lib_covr_dev
    ),
    .libPaths()
  )

  scenarios[[i, "covr version"]] <-
    as.character(packageVersion("covr", lib.loc = libs))

  x <- processx::process$new(
    command = file.path(R.home(), "bin", "R"),
    args = c(
      "--quiet",
      "--vanilla",
      "-e", paste0(collapse = " ", deparse(bquote(.libPaths(.(libs))))),
      "-e", "print(Sys.getenv()[which(startsWith(names(Sys.getenv()), 'REPREX'))])",
      "-e", "covr::package_coverage()"
    ),
    env = envvars,
    stdout = "|",
    stderr = "2>&1"
  )

  x$wait()
  scenarios[[i, "exit status"]] <- x$get_exit_status()
  cat(paste0("## ", x$read_all_output_lines(), collapse = "\n"), "\n\n")
}

scenarios
```

</details>

```{r, echo = FALSE, results = "asis"}
x <- scenarios
x[["exit status"]] <- ifelse(x[["exit status"]] == 0, ":white_check_mark:", ":x:")
x[["covr version"]] <- paste0("`v", x[["covr version"]], "`")

cat(
  sep = "",
  "| ", paste(colnames(x), collapse = " | "), " |\n",
  "| ", paste(rep_len("---", ncol(x)), collapse = " | "), " |\n",
  paste("|", apply(x, 1L, paste, collapse = " | "), "|", collapse = "\n"),
  "\n"
)
```

## Discussion

Documenting a few patterns I've noticed while trying to decipher this bug.
I'll walk through the table above from groups of scenarios from left to right.

### Fastmap Value

The value type stored is important. `covr` can safely handle integers, and
presumably other atomic data types without issue. I first encountered the
problem with `data.frame`s loaded using `parquet`, so that is what is used
otherwise.

From this, we can tell that solely _using_ fastmap to store values is not
sufficient to hit this bug.

## Fastmap Initialization

Next we can look at the way a `fastmap` object is initialized. For atomic
types, even a `fastmap` initialized in the package namespace will not through
an error. It's not entirely clear when `fastmap` might allocate memory, but I
assume this _is not_ something one should generally do on build anyways.

Scanning further down the table, we can see that using a `fastmap` defined in
the global environment is perfectly fine, so it is not sufficient to simply
use fastmap, it necessarily needs to be used within the package namespace.

Storing a `parquet` file in a simple list using `lazy-stub` (faking the
`fastmap` api) is also perfectly fine.

## `.onLoad` `if()`

We'll focus now on the last 12 rows, which showcase a preplexing behavior using
the CRAN version of `covr`.

Broadly, storing a `parquet` `data.frame` in a `fastmap` will cause a segfault,
whether the `fastmap` is initialized at buildtime, at runtime or lazily as the
`fastmap` is accessed by hooking in to the `$` generic.

_However_, bizarrely enough, there are certain snippets of code, that if
present somewhere in the package, will allow `covr` to succeed. One oddity that
I was able to pin down was whether or not `.onLoad` contained an `if` block. If
one exists (and in this case it's just a `if (TRUE) {}`), then `covr` is able
to dodge a segfault. I think this was also affected by whether or not an S3
generic was exported in the `NAMESPACE`, hence why S3 methods are registered
only `.onLoad`, but I didn't investigate this exhaustively.
